Glyph-maps for Visually Exploring Temporal Patterns in Climate Data and Models

### 2012-03-05 revision/paper.tex
minor changes from proofreading

The glyph-map uses a small glyph, or icon, to represent multiple variables, or time values, at each location. Glyph maps derive from glyph plots, which were initially developed to display multivariate data. One glyph is drawn for each observation in the data set. Types of glyph plots include the star plots \citep{mayr:1877, FD01, kleiner:1981}, the semi-graphic displays \citep{anderson:1960}, and the infamous Chernoff faces \citep{chernoff:1973}. An glyph is produced by mapping each variable to some graphical feature, such as the length of a line, curvature of the mouth, or size of the face.

@@

\begin{figure*}[htbp]
  \centering
  \includegraphics[width=5.5in]{nasa-colored-map.png}
  \caption{Facetted heatmap of de-seasonalized temperature. The dominant feature is the El Ni\~no warming in the southern equatorial region in the last half of 1997 and first half of 1998. Smaller features are only noticeable on closer inspection, or if pointed out: such as the relative warming on the mountain regions in south and north America which are red in all months in the later years.}
  \label{fig:nasa-facet}
\end{figure*}

@@

In general, multivariate glyph displays lack a natural ordering of the glyphs because observations have no ordering. Similarly, there are multiple ways that variables can be mapped to glyph property. It can result in a combinatorial explosion of possible mappings, each producing a display that may emphasize some entirely different features of the data than another mapping. This may have prevented the widespread adoption of glyph displays, despite some proposed ordering solutions \citep{kleiner:1981,hurley:2010}. Using glyphs with space-time data eliminates this issue: glyphs are placed according to the location of the measurement, and variables are ordered by the time they were collected. Additionally, climate data usually has high correlations between nearby locations and times. This imposes a degree of smoothness that gives the plot an appearance of a textured  landscape, making it easier to digest patterns from one glyph to another. Others \citep{pickett:1988} have used glyphs on maps to display multivariate spatial data. This area of work has morphed into the field of metaphorical data displays, which create abstract landscapes of spatial data, a digression from glyph-maps. \citet{gribov:2006} describes the use of glyph-maps for multivariate data also, with emphasis on the graphics software Gauguin. Glyph-maps for spatiotemporal data have been used in several publications \citep{carr:1992,eden:2010,hobbs:2010}. Glyph-maps using arrow glyphs have commonly been used for displaying ocean currents, wind direction and strength, as can be seen in the Figure 8.1 of \citet{IPCC} and a rudimentary time series glyph display can be seen in Figure 9.12.

@@

  \caption{Icon plots for 12 iconic time series shapes (linear increasing, decreasing, shifted, single peak, single dip, combined linear and nonlinear, seasonal trends with different scales, and a combined linear and seasonal trend) in Euclidean coordinates, time series icons (left) and polar coordinates, star plots (right). White reference grids, lines and circles are added to help compare the shapes.}
  \label{fig:templates}
\end{figure}

@@

This way of thinking of the glyph-map is convenenient for several reasons. The first is that it moves the plots beyond the small multiples, many little plots, approach. That provides flexibility to cover large spatial domains using potentially tiny glyphs. A second, and major, reason is that with this approach they can then be readily made to be interactive. Different variables can be swept into, and out of, the minor axes using mouse motion by varying $w$ and $h$ proportional to mouse movement. Interactive glyph-maps were used to explore the EXPO data using  GGobi \citep{swayne:2003}, and then later reproduced in publication quality resolution using {\tt ggplot2} \citep{me:ggplot2} in \citet{hobbs:2010}. In GGobi, a manual tour is used to make the linear combination of spatial variables with other variables \citep{CB95}, and equation \ref{coords.eqn} makes this approach more explicit. The third reason is that it enables adjusting scales of glyphs in multiple ways to vary the emphasis, which is discussed in Section \ref{sec:perception}.

@@

All plots facilitate some comparisons and impede others. For a plot to be useful for a particular data analytic task, primary comparisons need to be made the simplest by the display. For climate data, changes in slope, or trend, average value, and variance over time, are the primary tasks. Glyph-maps support these tasks because the pieces of the graphical elements that are needed to make the temporal comparisons are organized, and grouped together.

@@

Different features of the data can be emphasized by varying the scale used with each cell. By default, a global scale is used, so the same position within each cell corresponds to the same value in all locations. This facilitates the comparison of absolute values, and draws attention to plots with large overall variation.

@@

\begin{figure}[htbp]
 \centering
 \includegraphics[width=0.5\linewidth]{month-rescale01-col}%
 \includegraphics[width=0.5\linewidth]{month-rescale01-fill}
 \caption{Adding intensity and color to scaled plots. Here each location has been scaled to range $[0, 1]$ and color mapped to the range of the original predictions. (Left) Range mapped to intensity (grey scale) of the line: pale locations have smaller range, smaller effect, than darker locations, stronger effect. (Right) Range mapped to fill color of grid box: blue boxes have smaller ranges, and light boxes have larger effects. Shapes at all locations are visible, but attention is drawn to locations with lighter backgrounds, larger ranges, because of the higher contrast.}
 \label{fig:scaling-col}
\end{figure}

One more important point: if a model summary is displayed in the glyph-map, as discussed in the following section, it may be important to force the scale to that of the original data. The range of predicted values is typically smaller than the raw data, so angles of slopes is increased, giving an inaccurate sense of a stronger effect.

@@

The raw data shows us missing values and gives some sense of variability, but gives no information about long-term trend. Statistical models are useful for pre-processing climate data, to decompose data into long-term trend, seasonal effects and error, which is vital for making the problem both computationally and visually tractable. Trend is examined by fitting a linear model to each location and displaying predicted values in Figure~\ref{fig:gistemp-pred}. This plot shows long-term (de-seasonalized) temperature trend for 1950--2010 (the limited time period used due to missing data). Most locations have increasing temperature over this time period. Only a few locations show a decline, and these few locations tend to be different from their neighbors, and on the edge of the data collection region, indicating that there might be isolated data problems. The rate of increase differs across regions: steeper inclines in the north and western mountain regions and across the Great Lakes, and flatter inclines in the southeastern USA.

@@

  \caption{(Top) Glyph-map of predicted temperature anomaly data
    (GISTEMP), on reduced range temperature 1950-2010. Increasing
    trends can be seen over most of the USA, with varying degrees of
    increase. Some isolated locations exhibit steep declines, and
    inclines, perhaps indicative of some isolated data problems. Note
    that, the reference lines are straight, but they look curved
    because of the Z\"ollner optical illusion. (Bottom) Legend showing
    temperature and temporal ranges of individual glyphs.}
  \label{fig:gistemp-pred}
\end{figure}

@@

Regularly gridded spatial data helps with the perception of structure, because it helps to provide the reference frames upon which to make comparisons among the icons. When spatial locations are not on a regular grid, some difficulties arise. In this situation it is not clear what the scale of icons should be, there is no regular spacing. Some locations may be close to each other which would generate icons that overlap.

@@

\includegraphics[width=0.33\linewidth]{usa-lin-legend}
  %\includegraphics[width=0.25\linewidth]{ghcn-mountains}%
  \caption{(Top) Glyph-maps of the linear trend in ground station temperature values, 1950--2010, for USHCN stations, illustrating three issues for irregularly gridded data: overlapping icons, no natural choice for icon size, and the disordered appearance making comparisons more difficult. (Bottom) Legend showing
    temperature and temporal ranges of individual glyphs. }% (left) and stations in the mountain states (right).} 
  \label{fig:irregular}
\end{figure}
%The two panels illustrate one approach to the overlapping problem: zooming.  
One solution to the overlapping problem is to zoom in on smaller regions. There is an interplay between icon size (in geographical units), plotting area (in geographical units) and plot size (in display units).  Icons need to be big enough to be readable on the display, but small enough to avoid overlapping.  Readability can be maintained and overlapping reduced by decreasing the icon size and simultaneously increasing the display size or decreasing the plotting area.
% (as done in the right panel in Figure \ref{fig:irregular}).  
However, there are generally limitations in the maximum display size (or resolution) and the utility of plotting very small areas.

An alternative approach is to combine overlapping locations and plot a single summary icon.  In Figure \ref{fig:irregular-collapsed} nearby locations are collapsed to a single icon.  In the top panel each station still has an individual glyph but locations close to each other share a plotting area.  Not all glyphs are suited to this kind of space sharing, another option is to perform a summary of the locations that are collapsed and display that instead of the raw data.  The bottom panel shows an example of this where the average trend for the collapsed locations is displayed.  It is helpful to differentiate glyphs that involve summaries of more than one location, here a heavier line indicates at least two locations have been averaged.  The aggregation of spatial locations to a single summary can be misleading.  The sensitivity of analyses to the scale and configuration of aggregation is described by the  modifiable areal unit problem (\cite{Openshaw:1984kx, Fotheringham:1991uq}). In this type of glyph map the scale of aggregation is determined, rather arbitrarily, by the icon size. This dependence of visual inferences on the icon size is undesirable.  The ability to interactively modify icon size, and zoom quickly to disaggregate locations would go some way to alleviating the problem.

@@

In summary, glyph-maps enable the study of temporal patterns in multivariate spatiotemporal data. Climate change is focused on changes over time, so glyph-maps provide a way to study these changes directly. Glyph-maps enable different resolutions of the data to be examined: the raw data to discover data quality issues, early, global trend, seasonality, residuals, and multivariate dependencies. When the data is spatially gridded the glyphs are organized in a manner that makes reasonable comparisons. For irregularly gridded data the icon size needs to be chosen to minimize overlap but use as much display space as possible, or combined to produce close to gridded icons.

>>>>

The glyph-map uses a small glyph, or icon, to represent multiple variables, or time values, at each location. Glyph maps derive from glyph plots, which were initially developed to display multivariate data. One glyph is drawn for each observation in the data set. Types of glyph plots include the star plots \citep{mayr:1877, FD01, kleiner:1981}, the semi-graphic displays \citep{anderson:1960}, and the infamous Chernoff faces \citep{chernoff:1973}. A glyph is produced by mapping each variable to some graphical feature, such as the length of a line, curvature of the mouth, or size of the face.

@@

\begin{figure*}[htbp]
  \centering
  \includegraphics[width=5.5in]{nasa-colored-map.png}
  \caption{Facetted heatmap of de-seasonalized temperature. The dominant feature is the El Ni\~no warming in the southern equatorial region in the last half of 1997 and first half of 1998. Smaller features are only noticeable on closer inspection, or if pointed out: such as the relative warming on the mountain regions in South and North America which are red in all months in the later years.}
  \label{fig:nasa-facet}
\end{figure*}

@@

In general, multivariate glyph displays lack a natural ordering of the glyphs because the observations have no ordering. Similarly, there are multiple ways that variables can be mapped to glyph properties. It can result in a combinatorial explosion of possible mappings, each producing a display that may emphasize some entirely different features of the data compared to another mapping. This may have prevented the widespread adoption of glyph displays, despite some proposed ordering solutions \citep{kleiner:1981,hurley:2010}. Using glyphs with space-time data eliminates this issue: glyphs are placed according to the location of the measurement, and variables are ordered by the time they were collected. Additionally, climate data usually has high correlations between nearby locations and times. This imposes a degree of smoothness that gives the plot an appearance of a textured  landscape, making it easier to digest patterns from one glyph to another. Others \citep{pickett:1988} have used glyphs on maps to display multivariate spatial data. This area of work has morphed into the field of metaphorical data displays, which create abstract landscapes of spatial data, a digression from glyph-maps. \citet{gribov:2006} describes the use of glyph-maps for multivariate data also, with emphasis on the graphics software Gauguin. Glyph-maps for spatiotemporal data have been used in several publications \citep{carr:1992,eden:2010,hobbs:2010}. Glyph-maps using arrow glyphs have commonly been used for displaying ocean currents, wind direction and strength, as can be seen in the Figure 8.1 of \citet{IPCC} and a rudimentary time series glyph display can be seen in Figure 9.12.

@@

  \caption{Icon plots for 12 iconic time series shapes (linear increasing, decreasing, shifted, single peak, single dip, combined linear and nonlinear, seasonal trends with different scales, and a combined linear and seasonal trend) in Cartesian coordinates, time series icons (left) and polar coordinates, star plots (right). White reference grids, lines and circles are added to help compare the shapes.}
  \label{fig:templates}
\end{figure}

@@

This way of thinking of the glyph-map is convenient for several reasons. The first is that it moves the plots beyond the small multiples, many little plots, approach. That provides flexibility to cover large spatial domains using potentially tiny glyphs. A second, and major, reason is that with this approach they can then readily be made interactive. Different variables can be swept into, and out of, the minor axes using mouse motion, the scale can be adjusted by varying $w$ and $h$ proportional to mouse movement. Interactive glyph-maps were used to explore the EXPO data using  GGobi \citep{swayne:2003}, and then later reproduced in publication quality resolution using {\tt ggplot2} \citep{me:ggplot2} in \citet{hobbs:2010}. In GGobi, a manual tour is used to make the linear combination of spatial variables with other variables \citep{CB95}, and equation \ref{coords.eqn} makes this approach more explicit. The third reason is that it enables adjusting scales of glyphs in multiple ways to vary the emphasis, which is discussed in Section \ref{sec:perception}.

@@

All plots facilitate some comparisons and impede others. For a plot to be useful for a particular data analytic task, the display needs to make the primary comparisons the simplest to make. For climate data, changes in slope, or trend, average value, and variance over time, are the primary tasks. Glyph-maps support these tasks because the pieces of the graphical elements that are needed to make the temporal comparisons are organized, and grouped together.

@@

Different features of the data can be emphasized by varying the scale used within each cell. By default, a global scale is used, so the same position within each cell corresponds to the same value in all locations. This facilitates the comparison of absolute values, and draws attention to locations with large overall variation.

@@

\begin{figure}[htbp]
 \centering
 \includegraphics[width=0.5\linewidth]{month-rescale01-col}%
 \includegraphics[width=0.5\linewidth]{month-rescale01-fill}
 \caption{Adding intensity and color to scaled plots. Here each location has been scaled to range $[0, 1]$ and color mapped to the range of the original predictions. (Left) Range mapped to intensity (grey scale) of the line: pale locations have smaller effect, smaller range, than darker locations, stronger effect. (Right) Range mapped to fill color of grid box: blue boxes have smaller effects, and light boxes have larger effects. Shapes at all locations are visible, but attention is drawn to locations with lighter backgrounds, larger ranges, because of the higher contrast.}
 \label{fig:scaling-col}
\end{figure}

One more important point: if a model summary is displayed in the glyph-map, as discussed in the following section, it may be important to force the scale to that of the original data. The range of predicted values is typically smaller than the raw data, so angles of slopes are increased, giving an inaccurate sense of a stronger effect.

@@

The raw data shows us missing values and gives some sense of variability, but gives no information about long-term trend. Statistical models are useful for pre-processing climate data, to decompose data into long-term trend, seasonal effects and error, which is vital for making the problem both computationally and visually tractable. Trend is examined by fitting a linear model to each location and displaying predicted values in Figure~\ref{fig:gistemp-pred}. This plot shows long-term (de-seasonalized) temperature trends for 1950--2010 (the limited time period used due to missing data). Most locations have increasing temperature over this time period. Only a few locations show a decline, and these few locations tend to be different from their neighbors, and on the edge of the data collection region, indicating that there might be isolated data problems. The rate of increase differs across regions: steeper inclines in the north and western mountain regions and across the Great Lakes, and flatter inclines in the southeastern USA.

@@

  \caption{(Top) Glyph-map of predicted temperature anomaly data
    (GISTEMP), for the reduced period 1950-2010. Increasing
    trends can be seen over most of the USA, with varying degrees of
    increase. Some isolated locations exhibit steep declines, and
    inclines, perhaps indicative of some isolated data problems. Note
    that, the reference lines are straight, but they look curved
    because of the Z\"ollner optical illusion. (Bottom) Legend showing
    temperature and temporal ranges of individual glyphs.}
  \label{fig:gistemp-pred}
\end{figure}

@@

Regularly gridded spatial data helps with the perception of structure, because it helps to provide the reference frames upon which to make comparisons among the icons. When spatial locations are not on a regular grid, some difficulties arise. In this situation it is not clear what the scale of icons should be, as there is no regular spacing. Some locations may be close to each other which would generate icons that overlap.

@@

\includegraphics[width=0.33\linewidth]{usa-lin-legend}
  %\includegraphics[width=0.25\linewidth]{ghcn-mountains}%
  \caption{(Top) Glyph-maps of the linear trend in ground station temperature values, 1950--2010, for USHCN stations, illustrating three issues for irregularly gridded data: overlapping icons, no natural choice for icon size, and the disordered appearance making comparisons more difficult. (Bottom) Legend showing
    temperature and temporal ranges of individual glyphs. }% (left) and stations in the mountain states (right).} 
  \label{fig:irregular}
\end{figure}
%The two panels illustrate one approach to the overlapping problem: zooming.  
One solution to the overlapping problem is to zoom in on smaller regions. There is an interplay between icon size (in geographical units), area plotted (in geographical units) and plot size (in display units).  Icons need to be big enough to be readable on the display, but small enough to avoid overlapping.  Readability can be maintained and overlapping reduced by decreasing the icon size and simultaneously increasing the display size or decreasing the area plotted.
% (as done in the right panel in Figure \ref{fig:irregular}).  
However, there are generally limitations in the maximum display size (or resolution) and the utility of plotting very small areas.

An alternative approach is to combine overlapping locations and plot a single summary icon.  In Figure \ref{fig:irregular-collapsed} nearby locations are collapsed to a single icon.  In the left panel each station still has an individual glyph but locations close to each other share a plotting area.  Not all glyphs are suited to this kind of space sharing, another option is to perform a summary of the locations that are collapsed and display that instead of the raw data.  The right panel shows an example of this where the average trend for the collapsed locations is displayed.  It is helpful to differentiate glyphs that involve summaries of more than one location, here a heavier line indicates at least two locations have been averaged.  The aggregation of spatial locations to a single summary can be misleading.  The sensitivity of analyses to the scale and configuration of aggregation is described by the  modifiable areal unit problem (\cite{Openshaw:1984kx, Fotheringham:1991uq}). In this type of glyph map the scale of aggregation is determined, rather arbitrarily, by the icon size. This dependence of visual inferences on the icon size is undesirable.  The ability to interactively modify icon size, and zoom quickly to disaggregate locations would go some way to alleviating the problem.

@@

In summary, glyph-maps enable the study of temporal patterns in multivariate spatiotemporal data. Climate change is focused on changes over time, so glyph-maps provide a way to study these changes directly. Glyph-maps enable different resolutions of the data to be examined: the raw data to discover data quality issues, global trend, seasonality, residuals, and multivariate dependencies. When the data is spatially gridded the glyphs are organized in a manner that makes reasonable comparisons. For irregularly gridded data the icon size needs to be chosen to minimize overlap but use as much display space as possible, or combined to produce close to gridded icons.
